I would like to begin by thanking my principal supervisor Dr. Cathy Ramsbottom for her constant guifance and support throughout this work. Her reassurance and positivity has kept me going during my PhD studies. A special mention must also be made for Prof. Alan Hibbert, and Dr. Connor Ballance. Their constant discussions and perspectives on Physics has taught me more than I could have ever imagined.

It goes without saying that my experience at Queens has definitely been one of a kind, and I can only thank the rest of the staff members for everything that they do. I have also been to some great conferences and workshops around the world, and met some really amazing and influential scientists. In particular, I extend my thanks to Dr. Stuart Loch of Auburn University for his inspiration to some of this work.

I could not have asked for a better group of fellow PhD students and postdocs to share my experience with. I told myself that I would not name everybody but it would be a shame not to mention them. So thank you Simon, Steve, Sibasish, Matteo, Lorenzo, Ruari, Darren, Pietro, Marco, Andrew, Brendan, Kenny, Sam, Michael, Rian, Jack, Kathryn, Helena, Christine, and Luca, and those that have been and gone, Meredith, Stefan, David, and Mariona, Abel, and Jose. You have all made everyday working here unique and a pleasure regardless if Physics decides to work or not.

I leave this last paragraph for my family and friends. I would like to thank Hannah for her persistence and support through this thesis, I can only imagine what a burden is has been these past three years, (and arguably more!). My mum and dad (and Rolo of course) for their constant support, advice and absolutely everything else that has made me who I am today. The support of Kathryn, Martin, and James. Lisa Jamison and Sarah-Jane, and of course their lucky husbands David and Christ, and of course, Sally, Lily, and Annie. And of course all my other family members and friends, and collegues that are not mentioned that have made this hourney an unforgettable and unchangeable one.